---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First we globally set echo to TRUE.

```{r}
knitr::opts_chunk$set(echo=TRUE)
```

then we load some libraries which we will use later.

```{r}
library(dplyr)
library(lattice)
```

### 1. Loading the data
```{r}
if (!file.exists('activity.csv')) {
  unzip('activity.zip')
}
data <- read.csv('activity.csv', sep=',')
```

### 2. Pre-processing

We convert date strings to date objects. Reason: this will give better tick-labeling when plotting a time series.

```{r}
data$date <- as.POSIXct(data$date, format="%Y-%m-%d")
```

## What is mean total number of steps taken per day?

### 1. Total number of steps per day

Total per day is:
```{r}
total_data <- data %>% 
  group_by(date) %>%
  summarise(total = sum(steps, na.rm = TRUE))
with(total_data, plot(date, total, xlab='Date', ylab='Steps', main='Total number of steps per day'))
```

### 2. Histogram of totals per day

Let's first look at a histogram of the steps:
```{r}
with(total_data, hist(total, main='Histogram of steps', xlab='Steps', ylab='Frequency (days)'))
```

### 3. Calculate mean and median of totals per day

Mean of the totals per day is:
```{r}
with(total_data, mean(total))
```

Median of the totals per day is:
```{r}
with(total_data, median(total))
```


## What is the average daily activity pattern?

### 1. Time series plot of 5 min interval

```{r}
time_data <- data %>% 
  group_by(interval) %>%
  summarise(Mean = mean(steps, na.rm = TRUE))

with(time_data, plot(interval, Mean, type='l', xlab='Interval', ylab='Steps', main='Average number of steps per interval'))
```

### 2. Which interval contains max number of steps?

We now calculate the maximum of the mean value per interval and determine in what interval that maximum occurs.

```{r}
max_total <- with(time_data, max(Mean, na.rm = TRUE))
pos_max = with(time_data, which.max(Mean))
max_interval <- time_data[pos_max,]$interval
print(paste("Maximum number of average steps is ", max_total, ", which occurred in interval ", max_interval))
```

## Imputing missing values

### 1. Total number of records with  missing values

```{r}
print(paste("There are ", sum(is.na(data$steps)), " records with missing values"))
```


### 2. Filling in all of the missing values

Strategy: We use the *median per interval* to fill in missing values for that day.

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in

We now create a new dataset `imputed_data` with missing data filled in according to the strategy mentioned above.
```{r}
imputed_data <- data %>% 
  group_by(interval) %>% 
  mutate(steps = ifelse(is.na(steps), 
                        median(steps, na.rm = TRUE), 
                        steps))

print(paste("After imputation there are ", sum(is.na(imputed_data$steps)), " records with missing values"))

```

### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
total_imputed_data <- imputed_data %>% 
  group_by(date) %>%
  summarise(Sum = sum(steps))
```

Let's first look at a histogram of the steps:
```{r}
with(total_imputed_data, hist(Sum, main='Histogram of steps', xlab='Steps',    ylab='Frequency (days)'))
```

### 3. Calculate mean and median of totals per day

Mean of the totals per day is:
```{r}
with(total_imputed_data, mean(Sum))
```

Median of the totals per day is:
```{r}
with(total_imputed_data, median(Sum))
```

We see that the mean of the data with imputed values:

1. has become lower, and 
2. the median of the imputed data is the same as for the original data where we calculated them ignoring `NA` values.



## Are there differences in activity patterns between weekdays and weekends?


### 1. Create new factor variable weekday

We create a new factor 'weekday', which will get the value 'weekend' for Saturdays and Sundays, and 'weekday' for all other days.
```{r}
ndata <- imputed_data %>% 
  mutate(weekday=ifelse(weekdays(as.Date(date)) %in% c('Saturday', 'Sunday'), 'weekend', 'weekday'))
head(ndata)
```

### 2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

We split the data set in separate sets for weekend and weekdays and take the average for each interval.

```{r}
tdata <- ndata %>% 
  group_by(weekday, interval) %>%
  summarise(avg = mean(steps, na.rm = TRUE))

xyplot(avg ~ interval | factor(weekday), 
   data = tdata,
   ylab='Number of steps',
   xlab='Interval',
   main="Average number of steps per interval on weekdays and weekends",
   panel = function(x, y,...) {
           panel.xyplot(x, y,...)
           },
   type='l',
   layout=c(1,2)
   ) 

```


